---
title: "Helper Functions: Blending Multiple Waters"
# Blending water analysis using tidywater's helper functions
# Intro to Helper Functions Through a Blending Analysis
author: "Libby McKenna"
description: "This vignette will show how to use some of tidywater's helper functions through a water blending analysis."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{help_functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">#"
)
```

This vignette assumes a basic undertanding of `define_water` and the S4 `water` class. We'll need many `tidyverse` functions in addition to our `tidywater` functions for this analysis.

## Blending Analysis Setup
In this analysis, a hypothetical drinking water utility sources their water from a river and a lake, which both have high hardness. They are investigating whether blending up to 5 MGD from two groundwater wells will reduce the total hardness below 200 mg/L as CaCO3. 

## `define_water_once`
First, let's take a look at the available groundwater data from Well A and Well B. Notice the use of `define_water_once`.

```{r setup, warning=FALSE}
library(tidywater)
library(tidyverse)

# Example of how to use a tidywater "_once" function
data <- tibble(Well = c("A", "B"),
                ph = c(8, 9),
                alk = c(100, 150),
                temp = c(18, 19),
                ca_hard = c(5, 10),
                cond = c(500, 900),
                tds = c(300, 500),
                na = c(100, 200),
                k = c(0, 20),
                cl = c(0, 30),
                so4 = c(0,0)) %>%
  define_water_once()

data
```
`define_water_once` is the first helper function we'll cover. This function does exactly what `define_water` does, but applies it to a dataframe input. This means once your data has the proper column names, `define_water_once` can add the parameters to each `water` slot, calculate the carbonate balance (and other relevant calculations), and outputs all parameters in the `water` class as a dataframe. Any function with the `_once` suffix in `tidywater` can be used in a piped code block. However, `tidywater` functions cannot be used directly downstream of these types of functions because the data is no longer in a `water` class format. 

## `define_water_chain`
So what if you want to chain more tidywater functions together? In that case, you can use `define_water_chain`. This function takes a dataframe input, then outputs all parameters in a `water` class column. This is true for all `tidywater` functions with the `_chain` suffix. `_chain` functions are handy in a piped code block where you'll need to use many `tidywater` functions, such as `chemdose_ph`, `solvedose_alk`, etc. Most `tidywater` functions have a `_chain` or `_once` option.

```{r, warning=FALSE, echo=TRUE}

# Read in data from Wells A and B
raw_wells <- tibble(Well = c("A", "B"),
                    ph = c(8, 9),
                    alk = c(100, 150),
                    temp = c(18, 19),
                    ca_hard = c(5, 10),
                    cond = c(500, 900),
                    tds = c(300, 500),
                    na = c(100, 200),
                    k = c(0, 20),
                    cl = c(0, 30),
                    so4 = c(0,0)) %>%
  define_water_chain() %>% 
  balance_ions_chain()

raw_wells
```

Sometimes it's a good idea to verify our code is working properly. To make sure that our data was balanced using `balance_ions_chain`, we can plot our `water` class using `plot_ions`. The below example shows how to index a `water` class column: dataframe$water_class_column[[row_number]]


```{r, fig.width=7}
# Ion plot before balance_ions_chain was applied
raw_wells$defined_water[[1]] %>%
  plot_ions()
# Plot of balanced ions
raw_wells$balanced_water[[1]] %>%
  plot_ions()
```

Let's continue with our blending analysis. We're going to treat our two wells as a singular groundwater source. Blending can be calculated as Well_A_ratio * Well_A concentration + Well_B_ratio * Well_B_concentration. This is fine for most parameters, but for pH and acid/base equilibrium species, blending is a little more complicated. Enter: `blend_waters`. This function blends waters as you'd expect, and does all the pH blending math for you. In the example below, we're going to be blending inefficiently. But don't worry, there will be a better blending example later.

## `blend_waters`
To blend our two wells, we will blend row 1 of `balanced_water` with row 2 of `balanced_water`. This "vertical" blending is not efficient and will not be useful for large data frames. `water` objects cannot be pivoted, hence the row-to-row blending. In later examples, we will actually blend columns, which is more amenable to piped code chunks.
The `balanced_water` function takes 2 or more waters (must be of the `water` class), and corresponding ratios for each water. 

```{r, warning=FALSE}
 
# Blend "vertically": blends the data in well A's row with that of well B's. 
# The pluck function from the purrr package is useful for indexing a water class column
### First, index the water column using the name or number of the column (ie "balanced_water" or 3 (column number))
### Next, index the row

blended_wells <- blend_waters(waters = c(pluck(raw_wells, "balanced_water", 1),
                                         pluck(raw_wells, 3, 2)),
                              ratios = c(.5, .5))
# outputs a water class object.
blended_wells
```

To finish blending our wells, we will transform the `blended_wells` water object into a dataframe containing a `water` column. First, we'll use the `convert_water` function to to take the input `water` object and convert it to a dataframe. Then we'll run `define_water_chain` to make a dataframe with water column, named "wells". Most of `tidywater`'s `_chain` functions have the option to name the output column. Defaults vary depending on the `_chain` function. 

In addition to tidying the well water, we will also join all of the blend scenarios we will be modeling.

```{r, warning=FALSE}
# Assume wells can contribute 2.5 MGD each
groundwater <-   tibble(Wells_flow = c(0, 2.5, 5))
# Blending scenarios and the resulting source water ratios
scenarios <- tibble(surface_flow = seq(.1, 20, 0.1),
                    River_flow = c(seq(.1, 10, 0.1), rep(10, 100)),
                    Lake_flow = c(rep(0, 100), seq(.1, 10, .1)),
                    group = seq(1, 200, 1)) %>%
cross_join(groundwater) %>%
  mutate(total_flow = River_flow + Lake_flow + Wells_flow,
         River_ratio = River_flow/total_flow,
         Lake_ratio = Lake_flow/total_flow,
         Wells_ratio = Wells_flow/total_flow)

Wells <- convert_water(blended_wells) %>% 
  define_water_chain(output_water = "wells") %>% 
  select(-c(kw:tbaa)) %>% 
  cross_join(scenarios)
```

The river and lake sources don't require any mixing. We'll set up their raw data, balance the ions, and join in the blending scenarios.
```{r, warning=FALSE}

River <- tibble(ph = 7, temp = 20, alk = 200, tds = 950, cond = 1400,
                tot_hard = 300, na = 100, cl = 150, so4 = 200) %>%
  define_water_chain() %>%
  balance_ions_chain(output_water = "river") %>%
  cross_join(scenarios) %>%
  select(-defined_water)

Lake <- tibble(ph = 7.5, temp = 19, alk = 180, tds = 900, cond = 1000,
               tot_hard = 350, ca_hard = 250, na = 100, cl = 100, so4 = 150) %>%
  define_water_chain() %>%
  balance_ions_chain(output_water = "lake") %>%
  cross_join(scenarios) %>%
  select(-defined_water)
```

## `blend_waters_chain`
Now that we have our 3 sources defined, balanced, and cleaned up, we can blend them. This  next code chunk showcases the power of `tidywater.` We'll use `blend_waters_chain`, the helper function for `blend_waters`. We already created `water` class columns above, so we'll use those column names in the `waters` argument. The ratios for each water source were calculated in the `scenarios` dataframe. We'll pass the names of those ratio columns into the `ratio` argument. The ratios must always add up to 1, otherwise the function will not run.

```{r, warning=FALSE}
blend <- full_join(Wells, River) %>%
  full_join(Lake) %>%
  blend_waters_chain(waters = c("wells", "river", "lake"), ratios = c("Wells_ratio", "River_ratio", "Lake_ratio"))
```

## `pluck_water`
With all three source waters blended for each tested scenario, we can pull out a parameter of interest. In this case, we're investigating how much we can dilute the total hardness. This brings us to our final helper function for this vignette: `pluck_water`. This function uses `purrr::pluck` to create a new column for one selected parameter from a `water` class object. You can choose which `water` column to pluck from using the `input_water` argument. Next select the parameter of interest (which must match the water slot's name). Finally, the output column's name will default to the name of the plucked parameter, but there is an option to name it yourself using the `output_column` argument. If you want to pluck out more than one parameter, it is better to use a `_once` function instead of `_chain` (ie we could have used `blend_water_once` above).
```{r, fig.width= 7}
plotting <- blend %>% 
  pluck_water(input_water = "blended_water", "tot_hard") %>%
# Flag scenarios for plotting
    mutate(Flagged = case_when(tot_hard> 200 ~ "FLAG: Hardness > 200 mg/L CaCO3", TRUE ~ "Not Flagged"))

# Plot the results!
ggplot(plotting, aes(x = total_flow, y = tot_hard, color = as.character(Wells_flow), shape = Flagged))+
  geom_point()+
  scale_shape_manual(values = c(4, 16)) +
  labs(y = "Hardness (mg/L as CaCO3)", color = "Contributions from new wells (MGD)",
       shape = "Scenario Flags", x = "Total Plant Flow (MGD)") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.box = "vertical",
        legend.margin = margin(-5,0,0,0))
```

## Speed it up
As you use more `tidywater` helper functions with larger data sets, you'll notice the code can take a few minutes to run. All helper functions use functions from the <a href = "https://furrr.futureverse.org/"> furrr package</a>. To reduce processing time, you can activate the `furrr`'s parallel processing power by using `plan()` at the beginning of your script. `plan()` depends on what type of operating system you have, more info on that in the <a href = "https://furrr.futureverse.org/"> Controlling How Futures are Resolved</a> table. 

```{r, warning=FALSE}
library(furrr)

# For most operating systems, especially Windows, use this at the beginning of your script
plan(multisession)

#rest of script

```